Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-01-09 manuscript/manuscript.Rmd
updated with correlation explanation

Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations [@Clark2004; @Freckleton2011]. Such population-level data will never match the precision of individual-level data, but it is more feasible to attain a broad coverage sample when collecting coarse-scale data. This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predicitons but can be applied over greater spatial and temporal extents. An open question is how well models based on population-level data compare to models based on individual-level data.

To date, relatively few studies have tried to model populations based on data other than detailed individual-level data. An important exception is an effort by @Taylor2004 to model the population growth rate of an invasive species to investigate the best strategies for invasion control. They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure. Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 showed the application of such methods in a study on arable weeds. In particular, @Queenborough2011 provide emprical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models. Thus, population models based on coarse, population-level data show promise for producing ecological forecasts at landscape and regional scales [@Queenborough2011]. However, none of these models included environmental covariates.

@@

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therin for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September. The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment). In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions. The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program. Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.

@@

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit analagous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species. We describe the statistical models seprately since fitting the models required different approaches at the individual and quadrat levels. All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.

@@

We modeled growth as gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:

@@

where $x$ is log genet size and all other paramters are as described for the survival regression.

@@

We retain the abbreviations $S$, $G$, and $R$ from the analagous processes at the individual level. The vital rate models at the quadrat level are all based on quadrat proportional cover. Also, at the quadrat level we do not need to explicitly include a density dependent term. Since we are modeling proportional cover, we essentially get density-dependence for "free" when proportional cover is included as a covariate. That is, density-dependence emerges because quadrats with low cover gerenally increase in cover whereas quadrats with high cover generally decrease in cover.

@@

where, in this case, we do include random year effects on the intercept $\gamma$ and the slope term $\beta$. For cover change we had enough data for those terms to converge. Note that our model for quadrat cover change uses a logit transformation to link the expected cover at $t+1$ ($x_{jq,t}$) to the linear predictors. We do so because during model fitting we use a beta likelihood since the data, proportional cover, is beta distributed. The beta likelihood requires shape ($\rho$) and rate ($\eta$) parameters that can be caclulated using moment-matching:

@@

### Model fitting
Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those prior should be hierarchical. We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit heierarchically where species-specific coefficients were drawn from a shared `global' coefficient distribution. We did so for two reasons: (1) the four focal species are all perrennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strenght via partial pooling [@Gelman2007]. So, climate effects were modeled as:

@@

We used random draws from the final 1,000 iterations from each of 3 MCMC chains to introduce stochasticity into our population models. At each time step, we first randomly selected climate covariates from one of the 14 observed years and also randomly drew one set of random year effects for each vital rate regression. Then, we drew the full parameter set (specific random year effects, climate effects, fixed effects) from a randomly selected MCMC iteration. We selected random year effects and the climate year independently.

@@

Climate covariates improved the growth regressions at both levels, with a slghty greater contribution at the individual level (Figure 1). Parameter estimates were generally consistent in terms of sign and magnitude between the individual level and the quadrat level for each species and climate covariate (Figure 2). There is no clear pattern suggesting that parameters estimated using individual level data are more certain, as evidenced by credible interval widths, than estimates based on quadrat data.

@@

1. Climate detections
2. Limitations:
  * Sample size at quadrat level
  * are these species inherently sensitive to climate (compare to Chu and Adler in review -- say why we chose Montana site -- most sensitive to climate variables)
  * we know recruitment is important, and this is something that may be easily missed at quad level
3. Implications

>>>>

Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations [@Clark2004; @Freckleton2011]. Such population-level data will never match the precision of individual-level data, but it is more feasible to attain a broad coverage sample when collecting coarse-scale data. This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predictions but can be applied over greater spatial and temporal extents. An open question is how well models based on population-level data compare to models based on individual-level data.

To date, relatively few studies have tried to model populations based on data other than detailed individual-level data. An important exception is an effort by @Taylor2004 to model the population growth rate of an invasive species to investigate the best strategies for invasion control. They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure. Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 showed the application of such methods in a study on arable weeds. In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models. Thus, population models based on coarse, population-level data show promise for producing ecological forecasts at landscape and regional scales [@Queenborough2011]. However, none of these models included environmental covariates.

@@

Materials and Methods
---------------------
### Study site and data
Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therein for a complete description. The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September. The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).

From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment). In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions. The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program. Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.

@@

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit analogous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species. We describe the statistical models separately since fitting the models required different approaches at the individual and quadrat levels. All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.

@@

We modeled growth as Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:

@@

where $x$ is log genet size and all other parameters are as described for the survival regression.

@@

We retain the abbreviations $S$, $G$, and $R$ from the analogous processes at the individual level. The vital rate models at the quadrat level are all based on quadrat proportional cover. Also, at the quadrat level we do not need to explicitly include a density dependent term. Since we are modeling proportional cover, we essentially get density-dependence for "free" when proportional cover is included as a covariate. That is, density-dependence emerges because quadrats with low cover generally increase in cover whereas quadrats with high cover generally decrease in cover.

@@

where, in this case, we do include random year effects on the intercept $\gamma$ and the slope term $\beta$. For cover change we had enough data for those terms to converge. Note that our model for quadrat cover change uses a logit transformation to link the expected cover at $t+1$ ($x_{jq,t}$) to the linear predictors. We do so because during model fitting we use a beta likelihood since the data, proportional cover, is beta distributed. The beta likelihood requires shape ($\rho$) and rate ($\eta$) parameters that can be calculated using moment-matching:

@@

### Model fitting
Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those prior should be hierarchical. We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit hierarchically where species-specific coefficients were drawn from a shared `global' coefficient distribution. We did so for two reasons: (1) the four focal species are all perennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strength via partial pooling [@Gelman2007]. So, climate effects were modeled as:

@@

We used random draws from the final 1,000 iterations from each of 3 MCMC chains to introduce stochasticity into our population models. At each time step, we first randomly selected climate covariates from one of the 14 observed years and also randomly drew one set of random year effects for each vital rate regression. Climate years and random year effects were drawn independently because they were uncorrelated. Then, we drew the full parameter set (specific random year effects, climate effects, fixed effects) from a randomly selected MCMC iteration. We selected random year effects and the climate year independently.

@@

Climate covariates improved the growth regressions at both levels, with a slightly greater contribution at the individual level (Figure 1). Parameter estimates were generally consistent in terms of sign and magnitude between the individual level and the quadrat level for each species and climate covariate (Figure 2). There is no clear pattern suggesting that parameters estimated using individual level data are more certain, as evidenced by credible interval widths, than estimates based on quadrat data.

@@

1. Climate detection
2. Limitations:
  * Sample size at quadrat level
  * are these species inherently sensitive to climate (compare to Chu and Adler in review -- say why we chose Montana site -- most sensitive to climate variables)
  * we know recruitment is important, and this is something that may be easily missed at quad level
3. Implications
